Reformulating phylogenetic mixed models to improve flexibility and speed

### 2021-09-22 response.md

Why are the parameter estimates of the two implementations of the new model different in Fig. 5? And, in the same figure, why were runs of alternative implementations killed after 30 min?

**MLi: The parameters that are in both single and multiple site models remain the same. We have now added one more alternative implementation run for the 100 species simulation with the slow platforms. The goal of this paper is to show the inflexiably of existing tools and the efficiency of the new implemented approach. The implementation of glmmTMB designed to maximize speed and flexibility. All else equal, the computational performance of glmmTMB are expected to outperform lme4. We added a text in the results that cites this expectiation.**

I am not sure about the point raised in the Discussion: “Establishing the practical level of model complexity for a given problem and data set is an open and difficult general problem throughout statistical modeling, not just in phylogenetic studies.” Model complexity is typically addressed by model testing, which is not mentioned here.

**Todo: (for response and possibly add something to discussion) Problem 1: effects of model selection on coverage etc. (i.e. snooping), maybe not a problem if the aspects of the model being selected are not focal (e.g. Matuschek et al.); problem 2: understanding what levels of complexity are computationally practical**

@@

- In the same spirit of applied statistics, a brief description (or table) of which functions to use in the libraries lme4 and glmmTMB to make these models work would make it easier to start. There should also be mention of the tutorial available in the repository

**MLi: We have added an example in the vignette for this package**

- It would also be beneficial to add an application of the method with biological data directly in the article. The more your article is associated with biological ideas, the more users you will reach in my opinion.

**MLi: Thanks for the suggestion. This paper is written to be an application paper rather than a research paper. We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository (we could add this if the editor thought it was appropriate, but we're up against the length limits for an application paper and would prefer not to open the can of worms of making this into a research paper ...**

@@

**Thanks, these are wonderful and helpful suggestion. We have included a tutorial example in the released version on zenodo/github. We have now included the link in the paper.... **

Minor comments:

L26. In the Abstract, a claim has been made about the benefit of the method against unbalanced observational designs, but we lack insight on that point throughout the article. I think that sort of claim should be tested and I expect the Bayesian methods to perform better with that kind of problems.

**Table 1 identifies several broadly used packages that are restricted to balanced designs. (We are not discussing the performance (e.g. power or efficiency) of different methods applied to unbalanced designs, simply the fact that some software packages for PCM cannot handle unbalanced designs.)**

@@

**MLi: Fixed.**

Reviewer: 2

Comments to the Corresponding Author
This paper proposes an alternative formulation for the phylogenetic generalized linear mixed model which can utilize existing R packages for linear mixed models (lme4 and glmmTMB). The authors illustrate the advantages of their proposed method through simulations. The paper is well written and easy to follow. My main concern is that the proposed method is not scalable. That is, the computational efficiency of the proposed method only relies on the highly optimized implementation of lme4 and glmmTMB and its algorithm is actually "slower" than existing algorithms in phylogenetics. The followings are my specific comments:

**Todo: Why is it slower? Do we need to profile the pieces? Waffle!**

@@

2. In the paper, the authors replace the package MCMCglmm by the package brms. Although the implementation of MCMCglmm is not as efficient as brms, the complexity of MCMCglmm is linear with respect to the number of species while brms should be cubic (since it doesn't have a special method to avoid matrix inversion). It may be a good idea to add a simulation with a series of trees from dozens to thousands of species for showing the scalability of these methods (including MCMCglmm).

**MLi: Thanks. We have now added MCMCglmm to our simulation. For our simple simulation test, MCMCglmm is turned out to be faster based on your convergence critrion which heavily depend on the number of iteration. However, the goal is to look at the granlarity of frequentist approaches. Even for simple linear models, Bayesian approaches are much slowly due to MCMC compared to least squares.**

3. The models in simulations are not sufficiently described. What are the formulae for the phylogenetic random intercept variance, phylogenetic random slope variance, and covariance between phylogenetic random intercept and slope, and so on? Do they depend on the phylogenetic tree? The authors provide a clear explanation for how their proposed method can be applied to the standard phylogenetic regression (equation 1) but do not include it in the simulation section. I suggest the authors add this model to the simulation part for illustrating the applicability of their proposed method for this simple case.

**Todo: what? go back to the paper and see if we say this clearly, this should be trivial and clear. The formulas are clear! We are using the variance of the phylogenetic tree when simulating the b's.**

### 2021-09-24 response.md

As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).

**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application. We have provided an example in the vignette.**

The benchmarking of the new method is interesting but is mixing too many confounding factors linked with the different implementations of the various packages (see also comment from reviewer 2). What is the performance of the new model all else being equal? How much of the performance results is actually due to differences the optimization among the different R packages?

**MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency. The most direct comparison is between `phyloglmm` and the `phyr` or `pez` packages: all three methods are built on the mixed-models machinery of `lme4`, but `phyloglmm` modifies the Z matrix while `phyr` and `pez` modify the correlation structure of the random effects**. (TODO/MLi: is my restatement correct?)

@@

In the Introduction, a broader perspective about PCMs should be given before jumping into the specifics of the models tested here. Phylogenetic comparative methods include lots of other things, see e.g. Luke Harmon’s book: https://lukejharmon.github.io/pcm/chapters/. In general, a better reference to previous work is required and some claims of the paper should be either demonstrated or removed. For example, stating that “existing [PCM] procedures are either insufficiently flexible or too computationally demanding to analyze large data sets” is simply not true. There are many research papers applying PCMs to large datasets (and as a side note: how large should a dataset be to be considered large?).

**We have switched our focal set of models from "phylogenetic comparative methods" (PCM) to the narrower scope of "phylogenetic regression" (PR). The reviewer is correct that PCM is too broad a scope for describing our goals here. We demonstrate in our simulations that the fastest other methods in our set of comparisons take >30 minutes for a

@@

A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it. For example, Kostikova et al. 2016 (DOI:10.1093/sysbio/syw010) present a model that seems relevant here. Further, claiming that ‘brms’ is more efficient than ‘MCMCglmm’ needs to be backed up by either a reference or a test (see also comments from the reviewers).

** This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation and cited appropriately. **

In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.

**Bayesian methods are the state of the art for more complex models. I've now reworded the last paragraph in the intro before methods.**

### 2021-09-24 phyloglmm_ms_resubmit.tex
working on response

In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.

@@

We extend the simulation model by adding multiple groups where each group has one observation per species. 
The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
The full multi-group model is as follows: 
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix}
\right) \\
(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
\end{bmatrix}
\right) \\
\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
\label{eq:multiple_glmm}
\end{aligned}
\end{equation}
where $\textrm{I}_{\textrm{group}}$ is a indicator matrix for groups and $\otimes$ is the Kronecker product.

@@

We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\Sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
\bmb{we need a better colour for MCMCglmm.  colorspace package? }
    Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sstime.pdf}
  \caption{Comparison of single-group model computational speed.}
\label{ssplot_speed}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
  \caption{Comparison of coverage probability for fixed effect parameters. Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits. 
%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
  }
\label{ssplot_coverage}
\end{figure}
\end{center}

>>>>

In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets (e.g. random-slopes phylogenetic regressions with hundreds of species).
When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.

@@

We extend the simulation model by adding multiple groups where each group has one observation per species. 
The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
The full multi-group model is as follows: 
\begin{equation}
\begin{aligned}
\bY & = \dist(g^{-1}(\bmu), \phi) \\
\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int-slope}}} \\ 
\sigma_{\mathrm{phy_{int-slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix}
\right) \\
(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
\end{bmatrix}
\right) \\
\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
\label{eq:multiple_glmm}
\end{aligned}
\end{equation}
where $\textrm{I}_{\textrm{group}}$ is a indicator matrix for groups and $\otimes$ is the Kronecker product.

@@

We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\sigma_{\mathrm{phy_{int-slope}}}$).
Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.

@@

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
    Total simulations $N=100$ for each category. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
}
\label{ssplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sstime.pdf}
  \caption{Comparison of single-group model computational speed.}
\label{ssplot_speed}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
  \caption{Comparison of coverage probability for fixed effect parameters. Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits. 
%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
  }
\label{ssplot_coverage}
\end{figure}
\end{center}
